As the tech continues to evolve, software becomes an increasingly central part of our lives. From the apps on our phones to the software running our cars, it's everywhere. Everytime a  piece of code runs, it triggers energy usage inside the CPU, GPU, and other hardware components \cite{greenuk_footprint}. This subsequent usage of software adds up to the carbon emission a device generates \cite{thoughtworks_carbon_intensity}. The carbon footprint of software is a growing concern, as it contributes to climate change and has a significant impact on our planet. 

The scale of this concern is larger than most people realize. The information and communication technology (ICT) sector accounts for an estimated of 2.1\% to 3.9\% of the global greenhouse gas emissions \cite{ict_stat}. The global data center energy consumption alone is projected to reach around 945TWh by 2030, which is a significant increase from 153TWh in 2005, 205TWh in 2018.~\cite{iea_energy_consumption, masanet_energy_consumption}. Unlike the visible emissions from vehicles, software carbon is invisible. There is no smoke when a program runs, yet every computation contributes indirectly to the same atmosphere. This invisibility is precisely what makes it dangerous. It does not trigger the same urgency that the physical pollution does, even as the number grows.

One of the most consequential and least-studied of those decisions is the choice of programming language. The carbon cost of a running program is not determined by a hardware alone. It is also shaped by how efficient the language runtime executes instructions, manages memory, and utilizes the CPU.In a landmark empirical study, Pereira et al. \cite{10.1145/3136014.3136031} benchmarked 27 programming languages across 10 problems drawn from the Computer Language Benchmarks Game \cite{clbg}, measuring energy consumption via Intel's Running Average Power Limit (RAPL) interface. They found that the choice of programming language can have a significant impact on energy consumption, with some languages consuming up to 2.5 times more energy than others for the same task. For example, they found that C and Rust were among the most energy-efficient languages, while Python and JavaScript were among the least efficient. Python consumed approximately 75.88 times more energy per benchmark than C while JavaScript consumed approximately 4.45 times more energy per benchmark than C \cite{10.1145/3136014.3136031}. Since carbon emissions are directly proportional to energy consumption, these differences are not just technical concern, but also a carbon concern. A Python program consuming about 75 times more energy than an equivalent C program does not just run slower, it emits 75 times more carbon for every execution. Now, multiply that by the millions of times a piece of software runs, and the carbon cost becomes staggering.

Prior work in this area, including Pereira et al., reports energy consumption in joules or kilowatt-hours but none of it completes the translation into carbon footprint. This is a critical gap. Energy consumption is the underlying driver, but carbon footprint is the ultimate impact. This unit is most directly relevant to the climate policy and environmental decision making. Henderson et al. showed that the emissions can be reduced by up to 30 times just by running the experiments in the locations powered by more renewable energy sources \cite{10.5555/3455716.3455964}.

This paper presents a controlled empirical study comparing the energy consumption and the carbon footprint of six programming languages (C, Rust, Go, Java, Javascript, and Python) across six different algorithms and three input sizes. The energy consumption is measured using Intel's RAPL (accessed through LibreHardwareMonitor \cite{librehardwaremonitor}), and real-time carbon intensity data is fetched from the Electricity Maps API \cite{electricitymap} for the New York grid zone (US-NY-NYIS). Each of the 108 experimental cells (6 languages $\times$ 6 algorithms $\times$ 3 input sizes) was replicated 115 times, yielding 12,420 total measurements. Statistical analysis applies Welch's $t$-tests with Bonferroni correction across all 15 pairwise language comparisons. The results show that 14 of 15 language pairs are statistically distinguishable in carbon emissions. The sole exception is C versus Rust ($t = -0.313$, $p = 0.754$), confirming that the two languages are carbon-equivalent. Python is the most carbon-intensive language tested, producing on average $213\times$ more carbon emissions than C for identical workloads.