This section describes the experimental setup, the measurement protocol used to isolate algorithm energy consumption, the implementation constraints across all six languages, the input sizes used for each algorithm.

\subsection{Experimental Setup}
All 12,420 experiments were conducted on a single dedicated machine to eliminate any inter-machine power variability. 
The Intel's RAPL interface exposes hardware energy counters directly to software, enabling fine-grained CPU package energy measurement without external instruments 
\cite{10.1145/3177754}. 

\subsubsection{Implementation-specific design decisions}
All 36 implementations (6 languages x 6 algorithms) were written from scratch following a shared pseudocode specification. The following principles were followed to ensure a fair comparison across languages:

\paragraph{Algorithmic consistency -}All implementations strictly follow the same algorithmic logic and data structures as defined in the pseudocode specification. No language-specific optimizations or library functions that could alter the algorithm's behavior are used. Every algorithm is implemented from primitive operations only like loops, conditionals, integer arithmetic and array indexing.

\paragraph{Reset before each measurement -}Data structures that are mutated during  algorithm execution are reset to their initial state at the beginning of each algorithm invocation. Critically, this reset happens before the energy measurement window opens, so the reset cost is not included in the measured energy. This guarantees that every replicated run measures the same algorithmic work.

\paragraph{Output verification -}Each implementation produces a checksum output that is verified against a reference implementation to ensure correctness. This ensures that all implementations are performing the same computations and producing the same results.

\paragraph{Input size consistency -}The same input sizes are used for each algorithm across all languages to ensure that we are measuring comparable workloads. This allows us to analyze how energy consumption scales with input size for each algorithm and language.

\paragraph{Single checksum output -}Each implementation produces a single checksum output, and prints it to stdout after each algorithm execution. This is used for cross-language output verification and does not affect the measured energy as it occurs inside the measurement window but is identical and negligible across all languages.\\


\subsection{Stdin Measurement Protocol }

The main challenge in cross-language energy measurement is to ensure that we are measuring the energy consumed by the algorithm itself, and not the overhead of the language runtime or other background process. To address this, we designed a custom measurement harness that follows a strict protocol to isolate the algorithm execution. The harness uses a stdin/stout handshake design in which the program initializes once and then loops, executing the algorithm on each trigger without restarting the process.

The measurement sequence is shown in the Listing~\ref{lst:measurement_sequence}. 
\begin{lstlisting}[caption={Measurement sequence}, label={lst:measurement_sequence}, numbers=none]
    setup - not timed
    print "ready" - not timed
    harness newline - start timer
    run command - measurement
    print checksum - stop timer
    flush output - not timed
\end{lstlisting}

The flush() call on line 7 is critical. Without it, the newline sits in a kernel pipe buffer and the program never unblocks, causing the measurement to hang indefinitely. On the program side, each language blocks on its respective stdin/stdout primitive until the trigger arrives, executes the algorithm, and writes the checksum. Listing~\ref{lst:blocking_call} shows the blocking call used in each language.
\begin{lstlisting}[caption={Stdin primitives in each language implementation}, label={lst:blocking_call}, numbers=none]
    // C
    while (fgets(buf, sizeof(buf), stdin))  {run()}

    // Python
    for line in sys.stdin: run()

    // JavaScript 
    rl.on('line', () => run())

    // Java
    while(br.readLine() != null) {run()}

    // Go 
    for scanner.Scan() {run()}

    // Rust 
    for line in stdin.lock().lines() {run()}

\end{lstlisting}


C uses fgets rather than scanf because it reads exactly one line including the newline, giving unambiguous trigger semantics. Java uses BufferedReader.readLine() rather than Scanner to avoid locale-aware parsing overhead inside the measurement window. JavaScript uses Node.js's readline event listener. Python uses for line in sys.stdin: rather than repeated input() calls to avoid Python-level function call overhead per iteration.






\subsection{Experimental Execution}

\paragraph{Warm-up Runs -} Five runs per cell are discarded before measurement was recorded to allow the JIT compilers and garbage collectors in languages like Java and Go to warm up and reach a steady state. For C and Rust, warm-up primarily ensures that the instruction cache and branch predictor are in a warmed state consistent with repeated execution. 

\paragraph{Cell Sleep -} A mandatory 5s sleep is inserted between cells to allow the processor's thermal management to return to stable baseline between cells. This reduces the risk that thermal throttling in one cell biases the initial temperature state of the next.


\paragraph{Experiment timing -} All 36 language-algorithm-size combinations are executed in a single overnight run from 10pm to 6am to minimize background system load and to capture a consistent grid carbon intensity profile during the measurement window. 

\paragraph{Carbon Intensity Refresh -} The Electricity Maps API queried once at experiment start and then refreshed every 30 minutes. Two carbon intensity values were returned during the experiment window i.e. 300 gCO$_2$/kWh and 306 gCO$_2$/kWh, representing a 2\% spread in the carbon intensity during the experiment window. 




\begin{tcolorbox}[colback=gray!8, colframe=gray!40, 
                  title=\textbf{A Note on Background Interference}]
During development, an unplanned observation illustrated just how sensitive power-based energy measurement can be. While running early pilot iterations of Python matrix multiplication, energy  readings came in at 529--626~J per run, far above the expected  range. The cause turned out to be a YouTube video playing in the background on the same machine. Once the video was closed, readings dropped immediately to a stable 330--390~J. The CPU package power sensor captures everything the processor is doing, not just the algorithm being measured. This single observation shaped the entire experimental protocol: all measurements in the 
final study were collected overnight with no background applications running.
\end{tcolorbox}



\subsection{Output Verification}
Before energy measurement, all 36 implementations are verified for correctness. The verification script compiles and runs each implementation at all three input sizes, compares the checksum printed to stdout against the reference checksum produced by the C implementation, and assigns each language-algorithm-size combination a PASS/FAIL status. 
Table ~\ref{tab:c_run_verify} shows the verification results fot the C implemetation. The C times serves as a reference baseline against which all other languages are measured and normalized. They confirm that execution time scales with the expected time complexity of each program. 

\begin{table}[h]
    \centering
    \caption{C --- RUN \& VERIFY}\label{tab:c_run_verify}
    \small
    \begin{tabular}{lllll}
        \hline
        \textbf{Algorithm} & \textbf{Size} & \textbf{Output} & \textbf{Time} & \textbf{Status} \\
        \hline
            \texttt{summation}          & small    & 5050             &  0.0437ms  & PASS \\
            \texttt{summation}          & mid      & 50005000         &  0.0406ms  & PASS \\
            \texttt{summation}          & large    & 500000500000     &  0.4715ms  & PASS \\
            \texttt{binary\_search}     & small    & 99               &  0.0448ms  & PASS \\
            \texttt{binary\_search}     & mid      & 9999             &  0.0392ms  & PASS \\
            \texttt{binary\_search}     & large    & 999999           &  0.0835ms  & PASS \\
            \texttt{merge\_sort}        & small    & 100              &  0.0428ms  & PASS \\
            \texttt{merge\_sort}        & mid      & 10000            &  0.2459ms  & PASS \\
            \texttt{merge\_sort}        & large    & 1000000          & 24.7405ms  & PASS \\
            \texttt{bfs}                & small    & 100              &  0.0430ms  & PASS \\
            \texttt{bfs}                & mid      & 1000             &  0.3039ms  & PASS \\
            \texttt{bfs}                & large    & 5000             &  7.2774ms  & PASS \\
            \texttt{hash\_table}        & small    & 64850            &  0.0433ms  & PASS \\
            \texttt{hash\_table}        & mid      & 649985000        &  0.2467ms  & PASS \\
            \texttt{hash\_table}        & large    & 64999850000      &  2.1579ms  & PASS \\
            \texttt{matrix\_mul}        & small    & 9175             &  0.0744ms  & PASS \\
            \texttt{matrix\_mul}        & mid      & 646700           &  2.5604ms  & PASS \\
            \texttt{matrix\_mul}        & large    & 10291750         & 47.1134ms  & PASS \\
        \hline
    \end{tabular}
\end{table}

The execution times in Table~\ref{tab:c_run_verify} confirms the time complexity scaling. Merge sort grows from 0.0428ms to 24.740ms (a 576$\times$ increase) as input scales from small to large with consistent $O(n \log n)$ growth.


\subsection{Input Sizes}
Three input sizes are used for each algorithm to capture scaling behavior: small, mid, and large. The specific sizes are chosen to be large enough to produce measurable energy consumption while still completing within a reasonable time frame across all languages. The input sizes for each algorithm are as defined in Table~\ref{tab:input_sizes}.


\begin{table}[h]
    \centering
    \caption{Input Sizes per Algorithm}\label{tab:input_sizes}
    \small
    \begin{tabular}{lllll}
        \hline
        \textbf{Algorithm} & \textbf{Small} & \textbf{Mid} & \textbf{Large}   \\
        \hline
            \texttt{summation}          & 100       & 10,000    &  1,000,000    \\
            \texttt{binary\_search}     & 100       & 10,000    &  1,000,000    \\
            \texttt{merge\_sort}        & 100       & 10,000    &  1,000,000    \\
            \texttt{bfs}                & 100       & 1,000     &  5,000        \\
            \texttt{hash\_table}        & 100       & 10,000    &  100,000      \\
            \texttt{matrix\_mul}        & 50        & 200       &  500          \\
        \hline
    \end{tabular}
\end{table}




\subsection{Pilot Study and Sample Size Determination}
Before the main experiment, a pilot study of 50 iterations per cell was conducted to estimate per-cell variance and determine the required sample size \cite{Shechtman2013}. Per-cell coefficients of variation can be calculated with the following formula:
\begin{equation}
    CV = \frac{\sigma}{\mu} \times 100\%
    \label{eq:cv}
\end{equation}
Where $\sigma$ is the standard deviation of the energy measurements for a given cell, and $\mu$ is the mean energy measurement for that cell. The results revealed a wide and informative range. CV spanned from 4.4\% (Python matrix-multiplication, mid) to 122.7\% (JavaScript merge-sort, small), with an overall mean CV of 38.4\% and median of 43.0\%.
The high CV values observed at small input sizes across all languages are consistent with a known limitation of RAPL-based measurement. When an algorithm completes in microseconds, and OS scheduling event, cache miss, or background interrupt becomes a large fraction of the total measured energy, inflating variance \cite{10.1145/2425248.2425252}. This is clearly visible in our pilot data. The noisiest cells are uniformly those with short runtimes, while the most stable cells (Python matrix-multiplication, mid: CV=4.4\%, large: CV =6.9\% and Java merge-sort, large: CV=5.9\%) are those where the algorithm's own energy consumption is large enough to dominate the background noise, This pattern directly motivates the focus on large input size results in the main statistical analysis, where RAPL measurements are most reliable. 

A formal power analysis was then conducted using the \texttt{TTestIndPower} solver from the Python \texttt{statsmodels} library \cite{seabold2010} to determine the minimum sample size $N$ per cell required to detect a medium effect size (Cohen's $d = 0.5$) with 80\% power at the Bonferroni-corrected significance level of $\alpha' = 0.05/15 = 0.003\overline{3}$, using a two-sided Welch's $t$-test. The solver returned a raw $N$ of 116.28, which rounds up to a recommended minimum of $N = 117$. The main experiment used $N = 115$ per cell, 2 runs below the recommended minimum, a 1.7\% shortfall. Given the large effect sizes observed between paradigmatically different languages (particularly Python versus compiled languages), this marginal reduction in statistical power is negligible in practice and does not affect any of the study's conclusions. All 108 cells were replicated exactly 115 times, yielding 12,420 total measurements.



\subsection{Statistical Analysis}

All pairwise language comparisons are evaluated using \textbf{Welch's $t$-test} \cite{a967ba42-d0b9-3e65-976f-80cc6086b406} (also known as the unequal-variance $t$-test) rather than Student's $t$-test. Welch's test does not assume equal variance between the two groups being compared, making it more reliable when sample variances differ \cite{ruxton2006}. This is an important choice for this study: the six languages have substantially different variance in their gCO$_2$e distributions. Python's standard deviation is orders of magnitude larger than C's, making the equal-variance assumption of Student's $t$-test inappropriate.

With 6 languages, the total number of unique pairwise comparisons is $\binom{6}{2} = 15$. Running each comparison at a conventional significance level of $\alpha = 0.05$ would produce an unacceptably high family-wise false positive rate \cite{c3fa9fa7-dd2a-35f2-84a0-d5c07e68dd08}:
 
\begin{equation}
    P(\text{at least one false positive}) = 1 - (1 - 0.05)^{15} \approx 54\%
    \label{eq:fwer}
\end{equation}
 
To control this, the \textbf{Bonferroni correction}~\cite{c3fa9fa7-dd2a-35f2-84a0-d5c07e68dd08} is applied, dividing the nominal $\alpha$ by the number of simultaneous tests:
 
\begin{equation}
    \alpha' = \frac{\alpha}{m} = \frac{0.05}{15} = 0.003\overline{3}
    \label{eq:bonferroni}
\end{equation}
 
A pairwise comparison is declared statistically significant only if its $p$-value falls below $\alpha' = 0.0033$. This guarantees that the probability of any false positive across all 15 tests is controlled at 5\%.
 
Effect size is reported as \textbf{Cohen's $d$} \cite{cohen1988} for each significant pair, computed as the difference in group means divided by the pooled standard deviation:
 
\begin{equation}
    d = \frac{\bar{x}_A - \bar{x}_B}{s_{\text{pooled}}}, \qquad
    s_{\text{pooled}} = \sqrt{\frac{s_A^2 + s_B^2}{2}}
    \label{eq:cohend}
\end{equation}
 
By convention \cite{cohen1988}, $|d| < 0.2$ is negligible, $0.2 \leq |d| < 0.5$ is small, $0.5 \leq |d| < 0.8$ is medium, and $|d| \geq 0.8$ is large. Reporting effect size alongside $p$-values ensures that statistically significant differences are also practically meaningful, not merely artifacts of a large sample size \cite{cohen1988}.
 
All statistical tests are conducted on the large input size subset of the data, where energy measurements are most reliable and least affected by RAPL sensor noise at short durations \cite{10.1145/2425248.2425252}. At $N = 115$ measurements per cell, the Central Limit Theorem guarantees that the sampling distribution of the cell mean is approximately normal, satisfying the distributional assumption of the $t$-test without requiring explicit normality testing.